A generalization of the Bradley-Terry model for draws in chess

### 2019-05-15 inst/collusion_chess_draws.Rnw
work

In inference problems where the dataset is outcomes of paired
comparisons, such as chess, the Bradley-Terry model can be augmented
to chess-specific considerations in a number of ways.  Here I present
a new generalization of Bradley-Terry in which a chess match is
regarded as a three-way competition between the two players and an
entity that wins if the match is drawn.  Bradley-Terry can be further
modified to account for the White player advantage in a number of
ways.  This methodology affords insight into players' strengths,
response to playing Black or White, and risk-aversion as manifested by
probability of drawing.  The likelihood functions arising are easily
optimised numerically.  I analyse a number of datasets of chess
results, including the infamous World Chess Championships 1963 in
which the Soviets were accused of collusion.  I conclude that the
scoreline of the Candidates Tournament (Cura\c{c}ao 1962) exhibits no
evidence of collusion, in agreement with previous work.  However, I
find strong evidence of increased Soviet proclivity to draw in the
Interzonal Tournament, Stockholm 1962.

The likelihood functions arising are easily optimised numerically.  I
analyse a number of datasets of chess results, including the infamous
World Chess Championships 1963 in which the Soviets were accused of
collusion.

}

@@

Bradley-Terry is not directly applicable to Chess for two reasons:
firstly, the White player is at a considerable
advantage~\cite{asdfasdffdsfds}; and secondly, many chess matches end
in a draw.  Considering White advantage first, we find an overall mean
of about $(+0.375=0.394-0.276)$.  However, interpretations of this
statistic vary and the issue is hotly debated with opinions ranging
from Rowson who compares first-move advantage to that of serving in
tennis, to that of commentators such as \citeauthor{adorjan} who
claims that White's advantage is a ``delusion''; and that Black can
mount an intelligent and dynamic response to white: ``chess is a game
of complete information, and Black's information is always
greater---by one move!''.

@@

where $w_{ij}$ is the number of times $i$ beats $j$ and $l_{ij}$ is
the number of times $i$ loses to $j$.  Here I present a new
generalization of Bradley-Terry in which a chess match is regarded as
a three-way competition between the two players and an entity that
wins if the match is drawn.  The system is further modified to account
for first-move advantage by positing an entity whose strength is added
to that of the White player.  This methodology affords insight into
players' strengths, response to playing Black or White, and
risk-aversion as manifested by probability of drawing.

If we consider a chess match to be a three-way trial, we define a
likelihood function for non-negative $p_1,\ldots,p_n,p_D$ with
$p_D+\sum p_i=1$ and this would be

\begin{equation}
\prod_{i<j}
\left(\frac{p_i}{p_i+p_j+p_D}\right)^{W_{ij}}
\left(\frac{p_j}{p_i+p_j+p_D}\right)^{L_{ij}}
\left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}}
\end{equation}

where $W_{ij}$ is the number of games won by $i$ when playing $j$,
$L_{ij}$ is games lost and $D_{ij}$ is games drawn.  However, this
approach ignores the extra strength conferred by the first-move
advantage.  We thus consider another entity whose strength is added to
the white player.  The strength of the white player is thus the sum of
the player's strength and another entity, the White monster.
Recalling that a game might be a draw, we obtain the probabilities as
given in Table~\ref{probabilities}.  The corresponding likelihood
function is given below in equation \ref{likelihood_WD}:

@@

\begin{table}
\centering
\begin{tabular}{l|cc}\\
  &$i$ plays white & $j$ plays white\\ \hline
$\Prob{(\mbox{$i$ wins})}$  & $\frac{p_i+p_W\oo}{p_i+p_j+p_W+D}$ & $\frac{p_i}{p_i+p_j+p_W+D}$  \\
$\Prob{(\mbox{draw})}$      & $\frac{D\oo}{p_i+p_j+p_W+D}$       & $\frac{D}{p_i+p_j+p_W+D}$    \\
$\Prob{(\mbox{$i$ loses})}$ & $\frac{p_j\oo}{p_i+p_j+p_W+D}$     & $\frac{p_j+p_W}{p_i+p_j+p_W+D}$\\
\end{tabular}
\caption{Probabilities of a win, draw, loss for the two players under a generalized Bradley-Terry model\label{probabilities}}
\end{table}

\begin{align}\nonumber
\prod_{i<j}&{}
\left(\frac{p_i+p_W}{p_i+p_j+p_D+p_W}\right)^{W_{ij}^\mathrm{white}}
\left(\frac{p_j}{p_i+p_j+p_D+p_W    }\right)^{L_{ij}^\mathrm{white}}\\
&{} \left(\frac{p_i}{p_i+p_j+p_D+p_W}\right)^{W_{ij}^\mathrm{black}}\label{likelihood_WD}
\left(\frac{p_j+p_W}{p_i+p_j+p_D+p_W    }\right)^{L_{ij}^\mathrm{black}}\\
&{} \left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}}
\nonumber
\end{align}

@@

\begin{equation}
\prod_{i<j}
\left(\frac{p_D}{p_i+p_j+p_D+p_W}\right)^{D_{ij}^\mathrm{white}}
\left(\frac{p_D}{p_i+p_j+p_D+p_W}\right)^{D_{ij}^\mathrm{black}}
=
\prod_{i<j}
\left(\frac{p_D}{p_i+p_j+p_D}\right)^{D_{ij}^\mathrm{white}+D_{ij}^\mathrm{black}}
\end{equation}

which might be preferable on grounds of clarity.  We now apply this
and related likelihood models to real chess datasets.  The general
approach to testing any hypothesis $H$ on any likelihood function
$\mathcal{L}$ is first to identify the constraint on hypothesis space
corresponding to $H$.  One then finds the unconstrained maximum
likelihood $\mathcal{L}_\mathrm{free}$ and compares it with the
maximum likelihood $\mathcal{L}_\mathrm{cons}$ when the optimization
is constrained by $H$.  The test statistic is then the likelihood
ratio $\mathcal{L}_\mathrm{free}/\mathcal{L}_\mathrm{cons}$ or,
equivalently the support $\mathcal{S}=\log\mathcal{R}$.  Wilks's
theorem states that the asymptotic distribution of $2\mathcal{S}$ is
$\chi^2_n$ where $n$ is the number of degrees of freedom implemented
by $H$.

@@

The above likelihood functions are objective and reasonable
representations of chess match results.  However, they differ markedly
from previous likelihood functions of draws in that they include
entities representing white player advantage ($p_W$), and the
proclivity to draw $p_D$).   The $p_W$

@@

The result of numerical optimization is 7.304, that is, well over the
two-units-of-support per degree of freedom suggested by Edwards; we
can convincingly reject the hypothesis that $W=0$ and infer that
playing white confers a significant advantage.  Of course, this is
nothing new CITATION NEEDED, but it is now possible to go
significantly beyond mere existence of the effect and to quantify it.
Taking the MLE above, and considering a match between say Karpov and
Anand, the strengths are Karpov=0.064, Anand=0.151, white =0.104 and
draw =0.53.  Thus we would estimate that Karpov playing white against
Kasparov would win with probability
$\frac{0.064+0.104}{0.064+0.104+0.148+0.532}\simeq 0.198$, draw with
probability $\frac{532}{64+104+148+532}\simeq 0.625$, and lose with
probability $\simeq 0.174$ (compare $+0.075\,=0.627-0.297$ playing
black).

@@

might write $H_=\colon\mathbf{Kr}=\colon\mathbf{Ks}=\colon\mathbf{An}$

One very natural and interesting hypothesis would be that the three
players are of equal strength
%equal_strength_is_assessed_in_chunk_kka_equal_strength_above and we
might write $H_=\colon p_\mathrm{Kr}=p_\mathrm{Ks}=p_\mathrm{An}$.
This too is readily testable by the device of calculating the maximum
support over the region in which
$p_\mathrm{Kr}=p_\mathrm{Ks}=p_\mathrm{An}$.  In this case the extra
support in going from the constrained to the free optimization is
about 4.93.  However, this time the constraint corresponds to a loss
of {\em two} degrees of freedom so we would need 4 units of support,
which is attained in this case.  Alternatively we might follow Wilks,
and observe that the value of $2\mathcal{S}=2*4.92=9.84$ lies in the
tail region of its asymptotic distribution with a p-value of about
0.0072.  We can justifiably assert that the players do indeed have
different strengths, even if we account for white advantage and the
proclivity to draw.

@@

\subsection{Each player has a distinct White monster}

We now consider the possibility that the strength conferred by playing
White, is specific to each player and we denote this $w_\mathrm{An}$,
$w_\mathrm{Kr}$ and $w_\mathrm{Ks}$.  The corresponding likelihood
function would be

@@

In the likelihood function above we see, for example, that Anand won a
total of $18+6=24$ matches playing White, $13+2=15$ games playing
Black, and that Anand played white against Karpov a total of
$18+20+5=43$ times.  We fail to reject the null of
$w_\mathrm{An}=w_\mathrm{Kr}=w_\mathrm{Ks}$, having a likelihood ratio
of $4.5$, or an asymptotic $p$-value of $0.076$.  There is no strong
evidence that the players differ in regard to their white advantage.

@@

\noindent (the numerator includes 8 draws and 2 wins, one for Fischer
and one for Korchnoi, both playing White); but the full likelihood
function for the interzonal has 297 terms, including $D^{110}$.  To
investigate the possibility of Soviet collusion we introduce a new
draw entity: a soviet Draw monster $D_s$, who wins when two Soviet
players draw, and is inactive whenever a game with at least one
non-Soviet player is drawn.  Then $H_N\colon D_s=D$ corresponds to no
collusion and the likelihood function for table~\ref{interzonal} would
be

>>>>

In inference problems where the dataset is outcomes of paired
comparisons, such as chess, the Bradley-Terry model can be augmented
to chess-specific considerations in a number of ways.  Here I present
a new generalization of Bradley-Terry in which a chess match is
regarded as a three-way competition between the two players and an
entity that wins if the match is drawn.  Bradley-Terry can be further
modified to account for the white player advantage in a number of
ways.  Such reification affords insight into players' strengths,
response to playing black or white, and risk-aversion as manifested by
probability of drawing.  The likelihood functions arising are easily
optimised numerically.  I analyse a number of datasets of chess
results, including the infamous World Chess Championships 1963 in
which the Soviets were accused of collusion.  I conclude that the
scoreline of the Candidates Tournament (Cura\c{c}ao 1962) exhibits no
evidence of collusion, in agreement with previous work.  However, I
find strong evidence of increased Soviet proclivity to draw in the
Interzonal Tournament, Stockholm 1962.
}

@@

Bradley-Terry is not directly applicable to Chess for two reasons:
firstly, the white player is at a considerable
advantage~\cite{asdfasdffdsfds}; and secondly, many chess matches end
in a draw.  Considering white advantage first, we find an overall mean
of about $(+0.375=0.394-0.276)$.  However, interpretations of this
statistic vary and the issue is hotly debated with opinions ranging
from Rowson who compares first-move advantage to that of serving in
tennis, to that of commentators such as \citeauthor{adorjan} who
claims that white's advantage is a ``delusion''; and that Black can
mount an intelligent and dynamic response to white: ``chess is a game
of complete information, and Black's information is always
greater---by one move!''.

@@

where $w_{ij}$ is the number of times $i$ beats $j$ and $l_{ij}$ is
the number of times $i$ loses to $j$.  Here I present a new
generalization of Bradley-Terry in which a chess match is regarded as
a three-way competition between the two players and an entity that
wins if the match is drawn.  The system is further modified to account
for first-move advantage by positing an entity whose strength is added
to that of the white player.  This methodology affords insight into
players' strengths, response to playing black or white, and
risk-aversion as manifested by probability of drawing.

If we consider a chess match to be a three-way trial, we define a
likelihood function for non-negative $p_1,\ldots,p_n,D$ with
$D+\sum p_i=1$ and this would be

\begin{equation}
\prod_{i<j}
\left(\frac{p_i}{p_i+p_j+D}\right)^{W_{ij}}
\left(\frac{p_j}{p_i+p_j+D}\right)^{L_{ij}}
\left(\frac{D  }{p_i+p_j+D}\right)^{D_{ij}}
\end{equation}

where $W_{ij}$ is the number of games won by $i$ when playing $j$,
$L_{ij}$ is games lost and $D_{ij}$ is games drawn.  However, this
approach ignores the extra strength conferred by the first-move
advantage.  We thus consider another entity whose strength is added to
the white player.  The strength of the white player is thus the sum of
the player's strength and another entity, the White Monster.
Recalling that a game might be a draw, we obtain the probabilities as
given in Table~\ref{probabilities}.  The corresponding likelihood
function is given below in equation \ref{likelihood_WD}:

@@

\begin{table}
\centering
\begin{tabular}{l|cc}\\
  &$i$ plays white & $j$ plays white\\ \hline
$\Prob{(\mbox{$i$ wins})}$  & $\frac{p_i+W\oo}{p_i+p_j+W+D}$ & $\frac{p_i}{p_i+p_j+W+D}$  \\
$\Prob{(\mbox{draw})}$      & $\frac{D\oo}{p_i+p_j+W+D}$     & $\frac{D}{p_i+p_j+W+D}$    \\
$\Prob{(\mbox{$i$ loses})}$ & $\frac{p_j\oo}{p_i+p_j+W+D}$   & $\frac{p_j+W}{p_i+p_j+W+D}$\\
\end{tabular}
\caption{Probabilities of a win, draw, loss for the two players under a generalized Bradley-Terry model\label{probabilities}}
\end{table}

\begin{align}\nonumber
\prod_{i<j}&{}
\left(\frac{p_i+W}{p_i+p_j+D+W}  \right)^{W_{ij}^\mathrm{white}}
\left(\frac{p_j}{p_i+p_j+D+W    }\right)^{L_{ij}^\mathrm{white}}\\
&{} \left(\frac{p_i}{p_i+p_j+D+W}\right)^{W_{ij}^\mathrm{black}}\label{likelihood_WD}
\left(\frac{p_j+W}{p_i+p_j+D+W  }\right)^{L_{ij}^\mathrm{black}}\\
&{} \left(\frac{D}{p_i+p_j+D    }\right)^{D_{ij}}
\nonumber
\end{align}

@@

\begin{equation}
\prod_{i<j}
\left(\frac{D}{p_i+p_j+D+W}\right)^{D_{ij}^\mathrm{white}}
\left(\frac{D}{p_i+p_j+D+W}\right)^{D_{ij}^\mathrm{black}}
=
\prod_{i<j}
\left(\frac{p_D}{p_i+p_j+D}\right)^{D_{ij}^\mathrm{white}+D_{ij}^\mathrm{black}}
\end{equation}

which might be preferable on grounds of clarity.  We now apply this
and related likelihood models to real chess datasets.  The general
approach follows~\citet{edwards1972}.  To test any hypothesis $H$ on a
likelihood function $\mathcal{L}$ one first identifies the constraint
on hypothesis space corresponding to $H$.  One then finds the
unconstrained maximum likelihood $\mathcal{L}_\mathrm{free}$ and
compares it with the maximum likelihood $\mathcal{L}_\mathrm{cons}$
when the optimization is constrained by $H$.  The test statistic is
then the likelihood ratio
$\mathcal{L}_\mathrm{free}/\mathcal{L}_\mathrm{cons}$ or, equivalently
the support $\mathcal{S}=\log\mathcal{R}$.  Wilks's theorem states
that the asymptotic distribution of $2\mathcal{S}$ is $\chi^2_n$ where
$n$ is the number of degrees of freedom implemented by $H$.

@@

The above likelihood functions are objective and reasonable
representations of chess match results.  However, they differ markedly
from previous likelihood functions of draws in that they include
entities representing proclivity to draw $D$, and first-mover
advantage ($W$).

Considering white player a

The method of \citet{Edwards1972} involves only
numerical optimization (not integration which is computationally
expensive), and because gradients are available convergence is
generally rapid.

@@

The result of numerical optimization is 7.304, that is, well over the
two-units-of-support per degree of freedom suggested by
\citeauthor{edwards1972}; we can convincingly reject the hypothesis that
$W=0$ and infer that playing white confers a significant advantage.
This is consistent with expectation, but it is now possible to go
significantly beyond mere existence of the effect and to quantify it,
accounting for the differential strengths of the players.  Taking the
MLE above, and considering a match between say Karpov and Anand, the
strengths are Karpov=0.064, Anand=0.151, white =0.104 and draw =0.53.
Thus we would estimate that Karpov playing white against Kasparov
would win with probability
$\frac{0.064+0.104}{0.064+0.104+0.148+0.532}\simeq 0.198$, draw with
probability $\frac{532}{64+104+148+532}\simeq 0.625$, and lose with
probability $\simeq 0.174$ (compare $+0.075\,=0.627-0.297$ if Karpov
plays black).

@@

One very natural and interesting hypothesis would be that the three
players are of equal strength
%equal_strength_is_assessed_in_chunk_kka_equal_strength_above____________
and we might hypothesise that
$p_\mathrm{Kr}=p_\mathrm{Ks}=p_\mathrm{An}$.  This too is readily
testable by the device of maximizing the support, subject to the
constraint that $p_\mathrm{Kr}=p_\mathrm{Ks}=p_\mathrm{An}$.  In this
case the extra support in going from the constrained to the free
optimization is about 4.93, or a likelihood ratio of about $138$.
However, this time the constraint corresponds to a loss of {\em two}
degrees of freedom so we would need 4 units of support, which is
attained in this case.  Alternatively we might follow Wilks, and
observe that the value of $2\mathcal{S}=2*4.92=9.84$ lies in the tail
region of its asymptotic distribution with a p-value of about 0.0072.
We can justifiably assert that the players do indeed have different
strengths, even if we account for white advantage and the proclivity
to draw.

@@

\subsection{Each player has a distinct White Monster}

We now consider the possibility that the strength conferred by playing
white, is specific to each player and we denote this $w_\mathrm{An}$,
$w_\mathrm{Kr}$ and $w_\mathrm{Ks}$.  The corresponding likelihood
function would be

@@

In the likelihood function above we see, for example, that Anand won a
total of $18+6=24$ matches playing white, $13+2=15$ games playing
Black, and that Anand played white against Karpov a total of
$18+20+5=43$ times.  We fail to reject the null of
$w_\mathrm{An}=w_\mathrm{Kr}=w_\mathrm{Ks}$, having a likelihood ratio
of $4.5$, or an asymptotic $p$-value of $0.076$.  There is no strong
evidence that the players differ in regard to their white advantage.

@@

\noindent (the numerator includes 8 draws and 2 wins, one for Fischer
and one for Korchnoi, both playing white); but the full likelihood
function for the interzonal has 297 terms, including $D^{110}$.  To
investigate the possibility of Soviet collusion we introduce a new
draw entity: a soviet Draw monster $D_s$, who wins when two Soviet
players draw, and is inactive whenever a game with at least one
non-Soviet player is drawn.  Then $H_N\colon D_s=D$ corresponds to no
collusion and the likelihood function for table~\ref{interzonal} would
be
